SIMD\slash SIMT-vectorized Thermo-chemical Source Term Evaluation

### 2018-02-14 paper.tex
flesh out lit. review section

\subsection{Previous works and goals of this study}
Several recent works, e.g.~\cite{CurtisGPU:2017}, have investigated SIMT-based chemical kinetic integration on GPUs, however SIMD-based chemical kinetics have been less studied in comparison.
Stone et al., \cite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including CPUs, GPUs, and MICs.
The shallow-SIMD and SIMT-vectorization models were implemented in OpenCL and compared to an OpenMP baseline code that was deep SIMD-vectorized by simple compiler hints (a.k.a pragmas).
The shallow-SIMD vectorization improved the integrator performance over the OpenMP baseline by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was \SIrange{1.4}{1.6}{$\times$} slower than the OpenMP baseline due to thread-divergence concerns.
\cite{kroshko2013efficient} implemented a shallow-SIMD vectorized third order stiff Rosenbrock integrator for atmospheric chemistry on a Cell Broadband Engine---a specially designed vector processor---finding a speedup of \SI{1.89}{$\times$} over a serial version of the same code.

@@

\subsection{Code Generation and Testing Infrastructure}
Code generation is enabled by the python package \texttt{loo.py}~\cite{kloeckner_loopy_2014}, which translates user specified psuedo-code and data to OpenCL\slash C-code.
This allows for easy changes of program structure, e.g. data ordering, vectorization, threading patterns and easy switching of target-languages (and simpler extension to additional languages, e.g., CUDA).
Additionally, \texttt{loo.py} can execute developed subroutines from python (natively for C-code, and via \texttt{PyOpenCL}~\cite{kloeckner_pycuda_2012} for OpenCL-code), enabling unit-testing\slash validation for each component of the Jacobian or source terms.
The source-terms and subcomponents thereof (e.g., rates of progress, pressure-modification terms, etc.) are directly compared to Cantera~\cite{Cantera}, while the automatic differentiation code Adept~\cite{adept-v11} is used to obtain reference answers for Jacobian subcomponents.
Validation of the complete source-terms and Jacobian codes will be discussed in detail in~\cref{s:validation}.

\section{Results and Discussion}
\subsection{Source term validation}
\label{s:validation}
The reaction rates of progress (ROP), species and temperature rates in this study are validated by comparison with Cantera~\cite{Cantera}, however special care must be taken due floating point arithmetic issues.

@@

\subsection{Performance}
\label{S:results}
The performance studies in this work were run on a variety of different machines:

\begin{table}[htb]
\centering
\begin{tabular}{@{}l l l l l@{}}
\toprule
Model & Instruction Set & Vector Width & Cores & Identifier \\
\midrule
Intel Xeon X5650 & SSE4.2 & \SI{128}{\bit} & $2 \times 6$ & \texttt{sse4.2} \\
Intel E5-2690 V3 & AVX2 & \SI{256}{\bit} & $2 \times 12$ & \texttt{avx2} \\
\bottomrule
\end{tabular}
\label{t:cpus}
\caption{The CPU configurations used in this study.}
\end{table}
\begin{table}
\centering
\begin{tabular}{@{}l l l l l@{}}
\toprule
Model & NVIDIA Driver Version & CUDA Cores & Identifier \\
\midrule
Tesla C2075 & 384.81 & \num{448} & \texttt{C2075} \\
Tesla K40m & 375.26 & \num{2880} & \texttt{K40m} \\
\bottomrule
\end{tabular}
\label{t:gpus}
\caption{The GPU configurations used in this study.  NVIDIA's OpenCL runtime is provided with the graphic driver, rather than any specific version of CUDA.}
\end{table}

four ten-core \SI{2.2}{\giga\hertz} Intel Xeon E5-4640 v2 CPUs with \SI{20}{\mega\byte} of L3 cache memory, installed on an Ace Powerworks PW8027R-TRF+ with a Supermicro X9QR7-TF+/X9QRi-F+ baseboard.
Runtimes in each case were averaged over ten runs, each using the same set of PaSR conditions utilized in validation.
All code was compiled with \texttt{gcc 4.8.5} and ``\texttt{-O3}'' optimization, running on \texttt{v16.1.1} of the Intel OpenCL runtime supporting OpenCL \texttt{v1.2}.

>>>>

\subsection{Previous works and goals of this study}
A number of recent works have investigated the use of high-performance SIMT-devices to accelerate reactive-flow and chemical kinetics simulations.
Spafford et al.~\cite{Spafford:2010aa} investigated an implementation of an explicit direct numerical simulation code for compressible turbulent combustion on a Tesla C1060 GPU, and found an order of magnitude speedup in evaluation fo species production rates as compared to a serial CPU implementation on an AMD-Opteron processor.
Shi et al.~\cite{Shi:2011aa} developed a code to evaluate species rates and factorize the chemical kinetic Jacobian on a Tesla C2050 GPU, and found an order of magnitude speedup during integration of large chemical models when these operations were offloaded to the GPU as compared to the standard \texttt{CHEMKIN}~\cite{kee1989chemkin} and \texttt{LAPACK}~\cite{Anderson:1999aa} libraries on a quad-core Intel i7 930 CPU; it was not clear how\slash if the CPU code was parallelized.
Niemeyer et al.~\cite{Niemeyer:2011aa} implemented an explicit fourth-order Runge--Kutta integrator for non-stiff chemical kinetic integration on a Tesla C2075 GPU, and found a speedup of nearly two orders of magnitude when compared with a sequential CPU-code on an Intel Xeon X5650 CPU.
Shi et al.~\cite{Shi:2012aa} later extended their work to develop a stabilized-explicit solver on a Tesla C2050; paired with a traditional implicit integrator on quad-core Intel i7 930 which handled the stiffest computational cells, an \SIrange{11}{46}{$\times$} speedup was achieved over the base implicit CPU integrator during simulation of a three-dimensional premixed diesel engine simulation.
Le et al.~\cite{Le2013596} implemented two high-order shock-capturing reactive-flow codes on a Tesla C2070, and found a \numrange{30}{50}$\times$ speedup over a sequential CPU version of the same code on a Intel Xeon X5650.
Stone and Davis~\cite{Stone:2013aa} implemented the implicit VODE~\cite{Brown:1989vl} integrator on a Fermi M2050 GPU, achieving an order of magnitude speedup over the baseline CPU version on a single core of an AMD Opteron 6134 Magny-Cours.
Niemeyer and Sung~\cite{Niemeyer:2014aa} later developed a stabilized explicit second-order Runge--Kutta--Chebyshev algorithm on a Tesla C2075, demonstrating an order of magnitude speedup for a over a CPU implementation of VODE on a six-core Intel X5650 for moderately stiff chemical kinetics.
Sewerin and Rigopoulos~\cite{Sewerin20151375} implemented a three-stage\slash fifth-order implicit Runge--Kutta method~\cite{wanner1991solving} on a both high-end (Nvidia Quadro 6000) and consumer-grade (Nvidia Quadro 600) GPUs, as well as a standard CPU (two-core, four-thread Intel i5-520M) and a scientific workstation (eight-core, 16-thread Intel Xeon E5-2687W); the high-end GPU was at best \SI{1.8}{$\times$} slower than the workstation CPU (16 threads), while the consumer level GPU was at best \SI{5.5}{$\times$} slower than the corresponding standard CPU (four threads).
Niemeyer et al.~\cite{Niemeyer:2016aa} additionally developed and validated an analytical chemical kinetic Jacobian code for the CPU and GPU.
Yonkee and Sutherland~\cite{Yonkee2016} implemented SIMT-accelerated evaluations of thermodynamic parameters, multicomponent transport properties and species production rates for solution of a partial differential equations on an Intel Xeon E5-2680 CPU and an Tesla K20 GPU.
In evaluation of the thermochemical properties, speedups between \SIrange{8}{13}{$\times$} were on 16 CPU cores, and \SI{20}{40}{$\times$} on the GPU; speedups of \textasciitilde\SI{9}{$\times$} and \textasciitilde\SI{25}{$\times$} were found in solution of a partially premixed methanol flame PDE on 16 CPU cores and the GPU respectively.
Curtis et al.~\cite{CurtisGPU:2017} implemented a different formulation of the same fifth-order implicit Runge--Kutta method, as well as two-fourth order exponential integration techniques paired with an analytical Jacobian code~\cite{Niemeyer:2016aa} on a Tesla C2075; the implicit Runge--Kutta method was roughly equivalent to VODE running on 12--38 Intel Xeon E5-4640 v2 CPU cores for two smaller chemical models with an integration time step of \SI{e-6}{$\sec$}.

In contrast, SIMD-based chemical kinetics evaluation\slash integration have been far less studied.
Linford et al.~\cite{Linford:2011} implemented a three-stage, second-order Rosenbrock atmospheric chemical kinetics solver on the CPU, GPU and cell croadband engine (CBE)---a specially designed vector processor---and found speedups regularly exceeding~\SI{25}{$\times$} over a serial CPU implementation.
Kroshko and Spiteri~\cite{kroshko2013efficient} implemented a shallow-SIMD vectorized third order stiff Rosenbrock integrator for atmospheric chemistry on the CBE finding a speedup of \SI{1.89}{$\times$} over a serial version of the same code.
Stone et al.,~\cite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including CPUs, GPUs, and MICs.
The shallow-SIMD and SIMT-vectorization models were implemented in OpenCL and compared to an OpenMP baseline code that was deep SIMD-vectorized by simple compiler hints (a.k.a pragmas).
The shallow-SIMD vectorization improved the integrator performance over the OpenMP baseline by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was \SIrange{1.4}{1.6}{$\times$} slower than the OpenMP baseline due to thread-divergence concerns.

@@

\subsection{Code Generation and Testing Infrastructure}
Code generation is enabled by the python package \texttt{loo.py}~\cite{kloeckner_loopy_2014}, which translates user specified psuedo-code and data to OpenCL\slash C-code.
This allows for easy changes of program structure, e.g. data ordering, vectorization, threading patterns and easy switching of target-languages (and simpler extension to additional languages, e.g., CUDA).
Additionally, \texttt{loo.py} can execute developed subroutines from python (natively for C-code, and via \texttt{PyOpenCL}~\cite{kloeckner_pycuda_2012} for OpenCL-code), enabling unit-testing\slash validation for each component of the Jacobian or source terms.
The source-terms and subcomponents thereof (e.g., rates of progress, pressure-modification terms, etc.) are directly compared to Cantera~\cite{Cantera}, while the automatic differentiation code Adept~\cite{adept-v11} is used to obtain reference answers for Jacobian subcomponents.
The Portable OpenCL (POCL) implementation~\cite{poclIJPP} and OpenMP~\cite{dagum1998openmp} are used to run OpenCL and C unit-testing, respectively, on the continous integration framework Travis CI~\cite{travis:2018}.
Validation of the complete (as opposed to the subcomponent testing discussed here) generated source-terms and Jacobian codes will be discussed in detail in~\cref{s:validation}.

\section{Results and Discussion}
\subsection{Validation and performance study platforms}

The performance and validation studies for this work were run on a variety of CPU and GPU platforms.
\Cref{t:cpus} shows the number of cores, vector instruction set and model of the CPUs used in this work; each CPU had both \texttt{v16.1.1} of the Intel OpenCL runtime~\cite{intelopencl:2018} and \texttt{v1.0} of the POCL~\cite{poclIJPP} runtime installed, each enabling OpenCL \texttt{v1.2} execution.
In addition,~\cref{t:gpus} lists the model, number of CUDA cores and NVIDIA driver of each of the GPUs used in this work.
NVIDIA's OpenCL runtime is bundled with the NVIDIA driver~\cite{NVIDIA:2018}, hence we use driver version to specify the OpenCL runtime version.

\begin{table}[htb]
\centering
\begin{tabular}{@{}l l l l l@{}}
\toprule
CPU Model & Instruction Set & Vector Width & Cores & Identifier \\
\midrule
Xeon X5650 & SSE4.2 & \SI{128}{\bit} & $2 \times 6$ & \texttt{sse4.2} \\
E5-2690 V3 & AVX2 & \SI{256}{\bit} & $2 \times 12$ & \texttt{avx2} \\
\bottomrule
\end{tabular}
\label{t:cpus}
\caption{The Intel CPU configurations used in this study. The identifier field will be used as a shorthand descriptor in the performance plots to quickly identify the CPU type.}
\end{table}
\begin{table}
\centering
\begin{tabular}{@{}l l l l@{}}
\toprule
NVIDIA Model & Driver Version & CUDA Cores & Identifier \\
\midrule
Tesla C2075 & \num{384.81} & \num{448} & \texttt{C2075} \\
Tesla K40m & \num{375.26} & \num{2880} & \texttt{K40m} \\
\bottomrule
\end{tabular}
\label{t:gpus}
\caption{The NVIDIA GPU configurations used in this study.  NVIDIA's OpenCL runtime is provided with the graphic driver, rather than any specific version of CUDA.
Again, the identifier field will be used as a shorthand descriptor during analysis of the performance results.}
\end{table}

\subsection{Source term validation}
\label{s:validation}
The reaction rates of progress (ROP), species and temperature rates in this study are validated by comparison with Cantera~\cite{Cantera}, however special care must be taken due floating point arithmetic issues.

@@

\subsection{Performance}
\label{S:results}
The performance studies in this work were run on a variety of different machines.

four ten-core \SI{2.2}{\giga\hertz} Intel Xeon E5-4640 v2 CPUs with \SI{20}{\mega\byte} of L3 cache memory, installed on an Ace Powerworks PW8027R-TRF+ with a Supermicro X9QR7-TF+/X9QRi-F+ baseboard.
Runtimes in each case were averaged over ten runs, each using the same set of PaSR conditions utilized in validation.
All code was compiled with \texttt{gcc 4.8.5} and ``\texttt{-O3}'' optimization, running on  of the
